The Numerical Accuracy of Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.

### 2012-05-02 longRange.hetero/publish/sendfile/main.tex
fix bugs

The calculation of the long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction strength with respect to the distance between charges,
the typical short-range fast algorithms (cut-off + neighbor list
algorithms) do not work, and the
energy of the system is only conditionally convergent.  One of the
first algorithms handling this problem is the Ewald summation, dating
back to the 1920's~\cite{ewald1921die}.  It has been shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is  $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for modern
large scale molecular simulations.
Therefore, several Ewald based fast algorithms were proposed to reduce the
computational cost to a scalable level: $\mathcal O(N\log N)$. Some of them
are
the particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1}.
More recently, the staggered mesh or
``interlacing'' technique~\cite{chen1974reduction,eastwood1976optimal}
was applied to SPME
(called the staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called the interlaced P3M)~\cite{neelov2010interlaced}, and was showed to
improve the accuracy of these algorithms greatly.
In this paper, we focus on the Ewald summation, the SPME and
the staggered mesh Ewald methods. The Ewald summation is the starting
point of all the mentioned fast algorithm. The SPME
is the one of the most popular long-range algorithms among
the mainstream molecular simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a high dimensional parameter space.
The experiential choices may be either less accurate or waste
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one satisfying a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid bilayer membrane
solving in water: the hydrophobic tail(s) of the lipids carry no charge, and
the charges in the system
are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  Several
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this research is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
framework for the pairwise interaction is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.
% These error estimates are only of theoretical interest, and
% not applicable, because the computational costs of the error estimates
% are $\mathcal O(N^2)$ (homogeneity and inhomogeneity error)
% and $\mathcal O(N^3)$ (correlation error).
% If the error force has a special form so that the error estimates
% can be written as convolutions, the computational costs
% are reduced to $\mathcal O(N\log N)$ and 
% $\mathcal O(N^2\log N)$, respectively.
The computational cost of the homogeneity and inhomogeneity
error is $\mathcal O(N\log N)$ when the error force kernel is
of convolution form.
The  correlation error can also be
estimated at a cost of $\mathcal O(N\log N)$
by the nearest neighbor approximation
technique, which assume the error  is dominated by  short-range correlation.
This
is the case for most the liquid and gas systems far from critical point.
Generally speaking,
when the error force can be written as a pairwise form, 
all estimate are derived easily under this framework.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME method
and the staggered method Ewald method.   The estimates
of the homogeneity and inhomogeneous errors are tested by two
randomly distributed inhomogeneous charges systems,
% (i.e. the charges are not correlated),
and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

@@

certain neighboring range or charges connected by a chemical bonds.

@@

is important.  Therefore, in this case, we define the nearest neighbor by the

@@

\subsection{The smooth particle mesh Ewald method \&
the staggered mesh Ewald method}

@@

The SPME method provides two possibilities of calculating the
reciprocal force. The first one is to differentiate (with respect to
the particle position) the reciprocal term of the truncated Ewald
energy and then approximate the derived force \eqref{eqn:tmp8} by the
B-spline interpolation. This way is called
\emph{ik-differentiation}.
The alternative way notices the high-order
smoothness of the B-spline interpolation, and derive the force by
differentiating the B-spline approximated reciprocal Ewald energy
(also truncated). The second way was proposed by the original SPME paper
\cite{essmann1995spm}, and is called the \emph{analytical
  differentiation}.
It has been shown that using the same parameters,
the ik-differentiation is more precise than the analytical
differentiation, but it uses twice more FFTs, which is a bottleneck
for the communication in parallel
implementations~\cite{wang2010optimizing}.
For the ik-differentiation, the calculated force is not the negative
gradient of the calculated energy. The analytical differentiation
does not have this problem, but it (to the approximation precision)
violates the Newton's third law.
Since the SPME method is
not the point of the present paper, we refer the readers to, for
example, Ref. \cite{essmann1995spm, deserno1998mue1,
  wang2010optimizing} for details of this method.

@@

\subsection{Error estimate of the ik-differentiation}
\label{sec:error-ik}

@@

estimates, and leave all details of the derivation 

@@

$G_{\alpha,l}(\v r)$ does not depends on the coordinates of particles,

@@

\subsection{Error estimate of the analytical differentiation}
\label{sec:error-ana}

The error force of the analytical differentiation is 
\begin{align}\nonumber
  \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)
  % =\;&
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % e^{2\pi i\v m\cdot \v r}
  % \sum_{j\neq i}q_j\,
  % A(\v m, -\v r_j)
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % [A(\v m, \v r)\,
  % e^{2\pi i\v m\cdot \v r} ]
  % \sum_{j\neq i}q_j\,
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\\nonumber
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % A(-\v m, \v r)\,e^{2\pi i\v m\cdot \v r} \nabla_{\v r}e^{2\pi i\v m\cdot\v r} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % e^{-2\pi i\v m\cdot \v r}
  % \nabla_{\v r} [\,A(\v m, \v r) \,e^{2\pi i\v m\cdot\v r}\,]  \\\nonumber
  =&\,
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\\\nonumber
  & +\,
  q
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,
  \v B(\v m, \v r)\,e^{2\pi i\v m\cdot \v r}
  \sum_{j}q_j\,
  e^{-2\pi i\v m\cdot\v r_j} \\\label{eqn:ana.error.force}
  & +\,
  q^2
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,\v B(\v m, \v r) 
\end{align}
With $\v B(\v m, \v r)$ defined by
\begin{align}\nonumber
  \v B(\v m, \v r)
  &=
  \nabla_{\v r}\,A(\v m, \v r)\\\label{eqn:error-B}
  &=
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m)\,
  2\pi i l K_\alpha\v a_\alpha^\ast\,e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} 
\end{align}
The last term in Eqn.~\eqref{eqn:ana.error.force} is due to the
self-interaction~\cite{cerutti2009staggered, ballenegger2011removal,
  neelov2010interlaced}, which does not depends on the coordinates of
particles. Therefore, it can be calculated explicitly, and subtracted
from the analytical force during the simulation. The computational cost of
the self-interaction term is very low comparing with the SPME
force calculation. It is shown that the self-interaction  error
dominate when charge density is low~\cite{cerutti2009staggered}.
Therefore, we always remove the self-interaction term from
the analytical differentiation.
% , and call it as the original analytical
% differetiation when comparing it with the staggered mesh analytical
% differetiation.

@@

\subsection{Error estimate of the staggered mesh Ewald methods}

@@

We consider three inhomogeneous charge systems to verify our error
estimates.
The first two are artificially designed to verify the 
homogeneity and inhomogeneity error estimates. The third is a 
liquid-vapor equilibrium water system, in order to test the
quality  of error estimates in inhomogeneous and correlated system.

@@

\subsection{Example 1: a locally neutral system.}
\label{sec:example1}

@@

In this example,  41472 charges are put into a
$14.90\textsf{nm}\times 7.45\textsf{nm}\times 7.45\textsf{nm}$
periodic simulation box. One third of
them are carrying negative charge of $-0.834e$, and the other two thirds
are carrying positive charge of $+0.417e$.
The position of these charges
are randomly generated subjecting to the number density shown in
Fig.~\ref{fig:tmp-rho1}.  The number density of the  positive
charge   is twice as large as that of the negative charge,
so the
positive and negative charges cancel, and the
system is \emph{locally} neutral. Notice the amount of charges
and the density distributions are intentionally chosen the same as
Example 3~(see section \ref{sec:example3}), 
for an easy comparison.  The only difference is that the positions
of charges are independent  in this example,
while they are correlated in Example 3.
% Both the density distributions are highly
% inhomogeneous, take the positive distribution for example: between 11
% \textsf{nm} and 29 \textsf{nm}, the charge density is
% $1.0\;\textsf{e/nm}^3$, while between 0 \textsf{nm} to 9 \textsf{nm}
% and between 31 \textsf{nm} to 40 \textsf{nm}, the density is
% $0.05\;\textsf{e/nm}^3$.

@@

Fig.~\ref{fig:error1} presents the real RMS error (by points) and the
corresponding error estimates (by lines) of the cut-offed Ewald direct
part, the truncated Ewald reciprocal part, the original ik-/analytical
differentiation and the staggered mesh ik-/analytical
differentiation. All the error estimates are consistent very well
with the real errors.  The parameters are
chosen to be the same for an easy comparison among the
reciprocal space methods.  The errors of FFT
based fast Ewald methods are 8 orders of magnitudes large than that of
the truncated Ewald summation.
This supports the argument addressed in
Sec.~\ref{sec:error-ik}: the error introduced by truncating the
reciprocal Ewald summation is much smaller than that introduced by
approximating $e^{2\pi i\v m\cdot\v r}$.
The original ik-differentiation is more accurate than
the original analytical differentiation.
With the staggered mesh,
the error of the analytical differentiation reduces by more than 50\%,
while the accuracy of the ik-differentiation is only marginally
improved.
The errors of the staggered mesh ik- and analytical differentiation
are identical, which supports the theoretical prediction.
The original ik-differentiation uses 4
FFT transforms, and the original analytical differentiation uses
only 2.  Therefore, the staggered mesh analytical differentiation uses
the same amount of FFTs as the original ik-differentiation, and saves
one half FFTs than the staggered mesh ik-differentiation.  As mentioned before,
the FFT is a bottleneck of communication on massive parallel
super-computers,
so the methods using less FFTs may have advantages on these machines.

@@

\subsection{Example 2: separated positive and negative charge}
\label{sec:example2}

@@

Fig.~\ref{fig:error2} presents the real and estimated
RMS error of Example 2. The
the original ik-differentiation is not
shown, because it is nearly indistinguishable from the staggered
mesh ik-/analytical differentiation in this plot.
All the error estimates consist with the real errors.
Stemming from the inhomogeneity error,
two peaks of the direct error form at the interfacial
regions, and are much larger than the error in bulk
regions~\cite{wang2012}.
The reciprocal RMS errors only contain homogeneity contributions,
due to the varnished mean error force.
Similar with Example 1, the truncated Ewald
method is much more precise than the fast algorithms, and the
staggered mesh improves the accuracy of analytical
differentiation by more than 50\%.

@@

\subsection{Example 3: a water system in gas-liquid phase equilibrium}
\label{sec:example3}

@@

The real and estimated errors of the original SPME
(both ik- and analytical differentiation)
and the staggered mesh Ewald (both ik- and analytical differentiation)
are in Fig.~\ref{fig:water-error0}
and Fig.~\ref{fig:water-error}, respectively.
Unlike previous examples, the real error of ik- and analytical differentiation
are nearly the same.
The staggered mesh ik- and analytical differentiation
are also equivalent in this case, and are $\sim$70\% more accurate
than the original ik-/analytical differentiation.
In this case,
all estimates without considering the correlation error
deviate from the real errors.
The quality of the direct
part errors estimates is still good (though not perfect):
the error is overestimated by~50\%.
The RMS errors of the ik-differentiation, analytical differentiation
and the staggered mesh ik-/analytical differentiation are
overestimated by 2.6, 4.7 and 6.8 times, respectively.
Notice the only difference between this example
and Example 1 (in which the estimates were very sharp) is the correlation
of the charges in the system,
which plays a very important role in the RMS error. 
More surprisingly, the correlation
actually  reduces the error in the force computation.
Currently we have no explanation for this phenomena.
% Anyway, the correlation error $\mathcal
% E_{\textrm{correlation}}$ should be included to develop a more accurate error
% estimate. 
By including the nearest neighbor approximation of the
correlation error, the quality of all error estimates is improved greatly,
as indicated by the arrows in Fig.~\ref{fig:water-error0} and \ref{fig:water-error}.
Now the RMS errors of ik-differentiation, analytical differentiation
and staggered mesh ik-/analytical differentiation
are overestimated by a factor of 1.16, 1.44 and 1.86,
respectively.
Ref.~\cite{wang2010optimizing} showed that the error estimate
(ik-differentiation in a homogeneous water system) that
is even three times larger than the real error
still works well in the parameter tuning process, 
so the error estimate for  all methods are now good enough
for the parameter tuning application.
The great improvement by the nearest neighbor approximation 
means that  among all atomic
correlates (i.e. the rigid H-O bond and H-O-H angle, the van der Waals interactions,
the hydrogen bonds, etc.), the rigid bond and angle correlation
% (rigid
% modeling of H-O bond and H-O-H angle)
is the most important one, and all
the rest account for the remaining discrepancy between the estimated and real
error.

@@

Some minor contributions of the present work are:
\begin{itemize}\itemsep -3pt
\item In a locally neutral charge system, the inhomogeneity
  error was proved to vanish. Therefore, the error
  is of good property, and no adaptive method nor
  force correction~\cite{wang2012} is necessary. Fortunately, most
  molecular systems of practical interests are locally neutral.
\item We explicitly gave the expression of the self-interaction term
  in the analytical differentiation, which was proved to dominant the
  force error in low density systems~\cite{cerutti2009staggered}.  The
  expression helped removing the self-interaction term at a very low
  computational cost.
\item The error estimate showed that the staggered mesh Ewald
  (both ik- and analytical differentiation) are always more precise than
  the original SPME.
\item Unlike the original analytical differentiation
  that violates the Newton's third law, the staggered mesh analytical
  differentiation preserves it to the leading order.
\item We also proved the equivalence of the staggered mesh ik- and analytical
  differentiation, to the leading order of the force error.
  The staggered mesh analytical differentiation requires only one half
  FFTs as the ik-differentiation, so the former may be preferable
  for massive parallel simulation, due to the bottleneck of all-to-all
  communication required by the parallel FFT.
\end{itemize}

>>>>

The calculation of the long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction strength with respect to the distance between charges,
the typical short-range fast algorithms (e.g. the cut-off + neighbor list
algorithm) do not work, and the
energy of the system is only conditionally convergent.  One of the
first algorithms handling this problem is the Ewald summation, dating
back to the 1920's~\cite{ewald1921die}.  It has been shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is  $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for modern
large scale molecular simulations.
Therefore, several Ewald based fast algorithms were proposed to reduce the
computational cost to a scalable level: $\mathcal O(N\log N)$. Some of them
are
the particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1}.
More recently, the staggered mesh or
``interlacing'' technique~\cite{chen1974reduction,eastwood1976optimal}
was applied to SPME
(called the staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called the interlaced P3M)~\cite{neelov2010interlaced}, and was showed to
improve the accuracy of these algorithms greatly.
In this paper, we focus on the Ewald summation, the SPME and
the staggered mesh Ewald methods. The Ewald summation is the starting
point of all the mentioned fast algorithm. The SPME
is the one of the most popular long-range algorithms among
the mainstream molecular simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a 6-dimensional parameter space.
The experiential choices may be either less accurate or wasting
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one satisfying a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid bilayer membrane
solving in water: the hydrophobic tail(s) of the lipids carry no charge, and
the charges in the system
are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  Several
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this research is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
framework for the pairwise interaction is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.
% These error estimates are only of theoretical interest, and
% not applicable, because the computational costs of the error estimates
% are $\mathcal O(N^2)$ (homogeneity and inhomogeneity error)
% and $\mathcal O(N^3)$ (correlation error).
% If the error force has a special form so that the error estimates
% can be written as convolutions, the computational costs
% are reduced to $\mathcal O(N\log N)$ and 
% $\mathcal O(N^2\log N)$, respectively.
The computational cost of the homogeneity and inhomogeneity
error is $\mathcal O(N\log N)$ when the error force kernel is
of convolution form.
The  correlation error can also be
estimated at a cost of $\mathcal O(N\log N)$
by the nearest neighbor approximation
technique, which assume the error  is dominated by  short-range correlation.
This
is the case for most the liquid and gas systems far from critical point.
Generally speaking,
when the error force can be written as a pairwise form, 
all estimate are derived easily under this framework.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME 
and the staggered method Ewald methods.   The estimates
of the homogeneity and inhomogeneity errors are tested by two
randomly distributed inhomogeneous charges systems,
% (i.e. the charges are not correlated),
and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

  The first term on the r.h.s. of \eqref{eqn:error-split} is called
  the ``homogeneity error'', but it does NOT mean this term is
  homogeneous or originated from a homogeneous density profile.  It is
  named so because the error estimate only contains this term in the
  homogeneous and uncorrelated system. It is the term studied by most
  previous error estimates.

@@

certain neighboring range or charges connected by chemical bonds.

@@

is considered important.  Therefore, in this case, we define the nearest neighbor by the

@@

\subsection{The SPME \&
staggered mesh Ewald methods}

@@

The SPME method provides two possibilities of calculating the
reciprocal force. The first one is to differentiate (with respect to
the particle position) the reciprocal term of the truncated Ewald
energy and then approximate the derived force \eqref{eqn:tmp8} by the
B-spline interpolation. This way is called the
\emph{ik-differentiation}.
The alternative way notices the high-order
smoothness of the B-spline interpolation, and derive the force by
differentiating the B-spline approximated reciprocal Ewald energy
(also truncated). The second way was proposed by the original SPME paper
\cite{essmann1995spm}, and is called the \emph{analytical
  differentiation}.
It has been shown that using the same parameters,
the ik-differentiation is more precise than the analytical
differentiation, but it uses twice more FFTs, which is a bottleneck
for the communication in parallel
implementations~\cite{wang2010optimizing}.
For the ik-differentiation, the calculated force is not the negative
gradient of the calculated energy. The analytical differentiation
does not have this problem, but it (to the approximation precision)
violates the Newton's third law.
Since the SPME method is
not the point of the present paper, we refer the readers to, for
example, Ref. \cite{essmann1995spm, deserno1998mue1,
  wang2010optimizing} for details of this method.

@@

\subsection{Error estimate of the SPME ik-differentiation}
\label{sec:error-ik}

@@

estimate, and leave all details of the derivation 

@@

$\v G_{\alpha,l}(\v r)$ does not depends on the coordinates of particles,

@@

\subsection{Error estimate of the SPME analytical differentiation}
\label{sec:error-ana}

The error force of the analytical differentiation is 
\begin{align}\nonumber
  \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)
  % =\;&
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % e^{2\pi i\v m\cdot \v r}
  % \sum_{j\neq i}q_j\,
  % A(\v m, -\v r_j)
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % [A(\v m, \v r)\,
  % e^{2\pi i\v m\cdot \v r} ]
  % \sum_{j\neq i}q_j\,
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\\nonumber
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % A(-\v m, \v r)\,e^{2\pi i\v m\cdot \v r} \nabla_{\v r}e^{2\pi i\v m\cdot\v r} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % e^{-2\pi i\v m\cdot \v r}
  % \nabla_{\v r} [\,A(\v m, \v r) \,e^{2\pi i\v m\cdot\v r}\,]  \\\nonumber
  =&\,
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\\\nonumber
  & +\,
  q
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,
  \v B(\v m, \v r)\,e^{2\pi i\v m\cdot \v r}
  \sum_{j}q_j\,
  e^{-2\pi i\v m\cdot\v r_j} \\\label{eqn:ana.error.force}
  & +\,
  q^2
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,\v B(\v m, \v r),
\end{align}
with $\v B(\v m, \v r)$ defined by
\begin{align}\nonumber
  \v B(\v m, \v r)
  &=
  \nabla_{\v r}\,A(\v m, \v r)\\\label{eqn:error-B}
  &=
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m)\,
  2\pi i l K_\alpha\v a_\alpha^\ast\,e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r}.
\end{align}
The last term in Eqn.~\eqref{eqn:ana.error.force} is due to the
self-interaction~\cite{cerutti2009staggered, ballenegger2011removal,
  neelov2010interlaced}, which does not depends on the coordinates of
particles. Therefore, it can be calculated explicitly, and subtracted
from the analytical force during the simulation. The computational cost of
the self-interaction term is low comparing with the SPME
force calculation. It is shown that the self-interaction  error
dominate when charge density is low~\cite{cerutti2009staggered}.
Therefore, we always remove the self-interaction term from
the analytical differentiation.
% , and call it as the original analytical
% differetiation when comparing it with the staggered mesh analytical
% differetiation.

@@

\subsection{Error estimate of the staggered mesh Ewald method}

@@

We consider three inhomogeneous charge systems to verify our error
estimates.
The first two are artificially designed to verify the 
homogeneity and inhomogeneity error estimates. The third is a 
liquid-vapor equilibrium water system, in order to test the
quality  of the nearest neighbor approximation for the correlated system.

@@

\subsection{Example 1: A locally neutral inhomogeneous system}
\label{sec:example1}

@@

In this example,  41472 charges are put into a
$14.90\textsf{nm}\times 7.45\textsf{nm}\times 7.45\textsf{nm}$
periodic simulation box. One third of
them are carrying negative charge of $-0.834e$, and the other two thirds
are carrying positive charge of $+0.417e$.
The position of these charges
are randomly generated subjecting to the number density shown in
Fig.~\ref{fig:tmp-rho1}.  The number density of the  positive
charge   is twice as larger as that of the negative charge,
so the
positive and negative charges cancel, and the
system is \emph{locally} neutral. Notice the amount of charges
and the density distributions are intentionally chosen the same as
Example 3~(see section \ref{sec:example3}), 
for an easy comparison.  The only difference is that the positions
of charges are independent  in this example,
while they are correlated in Example 3.
% Both the density distributions are highly
% inhomogeneous, take the positive distribution for example: between 11
% \textsf{nm} and 29 \textsf{nm}, the charge density is
% $1.0\;\textsf{e/nm}^3$, while between 0 \textsf{nm} to 9 \textsf{nm}
% and between 31 \textsf{nm} to 40 \textsf{nm}, the density is
% $0.05\;\textsf{e/nm}^3$.

@@

Fig.~\ref{fig:error1} presents the real RMS error (by points) and the
corresponding error estimates (by lines) of the cut-offed Ewald direct
part, the truncated Ewald reciprocal part, the original ik-/analytical
differentiation and the staggered mesh ik-/analytical
differentiation. All the error estimates are consistent very well
with the real errors.  The parameters are
chosen to be the same for an easy comparison among the
reciprocal space methods.  The errors of FFT
based fast Ewald methods are 8 orders of magnitudes larger than that of
the truncated Ewald summation.
This supports the argument addressed in
Sec.~\ref{sec:error-ik}: the error introduced by truncating the
reciprocal Ewald summation is much smaller than that introduced by
approximating $e^{2\pi i\v m\cdot\v r}$.
The original ik-differentiation is more accurate than
the original analytical differentiation.
With the staggered mesh,
the error of the analytical differentiation reduces by more than 50\%,
while the accuracy of the ik-differentiation is only marginally
improved.
The errors of the staggered mesh ik- and analytical differentiation
are identical, which supports the theoretical prediction.
The original ik-differentiation uses 4
FFT transforms, and the original analytical differentiation uses
only 2.  Therefore, the staggered mesh analytical differentiation uses
the same amount of FFTs as the original ik-differentiation, and saves
one half FFTs than the staggered mesh ik-differentiation.  As mentioned before,
the FFT is a bottleneck of communication on massive parallel
super-computers,
so the methods using less FFTs may have advantages on these machines.

@@

\subsection{Example 2: Separated positive and negative charge}
\label{sec:example2}

@@

Fig.~\ref{fig:error2} presents the real and estimated
RMS error of Example 2. The
the original ik-differentiation is not
shown, because it is nearly indistinguishable from the staggered
mesh ik-/analytical differentiation in this plot.
All the error estimates consist with the real errors.
Stemming from the inhomogeneity error,
two peaks of the direct error form at the interfacial
regions, and are much larger than the error in bulk
regions.
The reciprocal RMS errors only contain homogeneity contributions,
due to the varnished mean error force.
Similar with Example 1, the truncated Ewald
method is much more precise than the fast algorithms, and the
staggered mesh improves the accuracy of analytical
differentiation by more than 50\%.

@@

\subsection{Example 3: A water system in gas-liquid phase equilibrium}
\label{sec:example3}

@@

The real and estimated errors of the original SPME
(both ik- and analytical differentiation)
and the staggered mesh Ewald (both ik- and analytical differentiation)
are in Fig.~\ref{fig:water-error0}
and Fig.~\ref{fig:water-error}, respectively.
Unlike previous examples, the real error of ik- and analytical differentiation
are nearly the same.
The staggered mesh ik- and analytical differentiation
are also equivalent in this case, and are $\sim$70\% more accurate
than the original ik-/analytical differentiation.
In this case,
all estimates without considering the correlation error
deviate from the real errors.
The quality of the direct
part error estimate is still good (though not perfect):
the error is overestimated by~50\%.
The RMS errors of the ik-differentiation, analytical differentiation
and the staggered mesh ik-/analytical differentiation are
overestimated by 2.6, 4.7 and 6.8 times, respectively.
Notice the only difference between this example
and Example 1 (in which the estimates were very sharp) is the correlation
of the charges in the system,
which plays a very important role in the RMS error. 
More surprisingly, the correlation
actually  reduces the error in the force computation.
Currently we have no explanation for this phenomena.
% Anyway, the correlation error $\mathcal
% E_{\textrm{correlation}}$ should be included to develop a more accurate error
% estimate. 
By including the nearest neighbor approximation of the
correlation error, the quality of all error estimates is improved greatly,
as indicated by the arrows in Fig.~\ref{fig:water-error0} and \ref{fig:water-error}.
Now the RMS errors of ik-differentiation, analytical differentiation
and staggered mesh ik-/analytical differentiation
are overestimated by a factor of 1.16, 1.44 and 1.86,
respectively.
Ref.~\cite{wang2010optimizing} showed that the error estimate
(ik-differentiation in a homogeneous water system) that
is even three times larger than the real error
still works well in the parameter tuning process, 
so the error estimate for  all methods are now good enough
for the parameter tuning application.
The great improvement by the nearest neighbor approximation 
means that  among all atomic
correlates (i.e. the rigid H-O bond and H-O-H angle, the van der Waals interactions,
the hydrogen bonds, etc.), the rigid bond and angle correlation
% (rigid
% modeling of H-O bond and H-O-H angle)
is the most important one, and all
the rest account for the remaining discrepancy between the estimated and real
error.

@@

Some minor contributions of the present work are:
\begin{itemize}\itemsep -3pt
\item In a locally neutral charge system, the inhomogeneity
  error was proved to vanish. Therefore, the error
  is of good property, and no adaptive method nor
  force correction~\cite{wang2012} is necessary. Fortunately, most
  molecular systems of practical interests are locally neutral.
\item We explicitly gave the expression of the self-interaction term
  in the analytical differentiation, which was proved to dominant the
  force error in low density systems~\cite{cerutti2009staggered}.  The
  expression helped removing the self-interaction term at a comparatively low
  computational cost.
\item The error estimate showed that the staggered mesh Ewald
  (both ik- and analytical differentiation) are always more precise than
  the original SPME.
\item Unlike the original analytical differentiation
  that violates the Newton's third law, the staggered mesh analytical
  differentiation preserves it to the leading order.
\item We also proved the equivalence of the staggered mesh ik- and analytical
  differentiation, to the leading order of the force error.
  The staggered mesh analytical differentiation requires only one half
  FFTs as the ik-differentiation, so the former may be preferable
  for massive parallel simulation, due to the bottleneck of all-to-all
  communication required by the parallel FFT.
\end{itemize}
